# locationallocation <a href = "https://giacfalk.github.io/locationallocation/"><img src = "man/figures/logo.png" align="right" width="120" /></a>

<!-- quarto render -->

<!-- Install the package before rendering this file: `devtools::install()` -->

```{r}
#| label: setup
#| include: false

library(locationallocation)

library(badger)
library(ggplot2)
library(here)
library(magrittr)
library(sf)
library(stars)

source(here("R", ".setup.R"))
```

<!-- badges: start -->
```{r}
#| echo: false
#| output: asis
#| message: false
#| warning: false

cat(
  badge_repostatus("active"),
  badge_custom(
    x = "doi",
    y = "10.31223/X5XQ69",
    color = "1284C5",
    url = "https://doi.org/10.31223/X5XQ69"
  ),
  badge_github_actions(action = "R-CMD-check.yaml"),
  badge_license(
    license = "GPLv3",
    color = "bd0000",
    url = "https://www.gnu.org/licenses/gpl-3.0"
  ),
  badge_custom(
    x = "Contributor%20Covenant",
    y = "3.0",
    color = "4baaaa",
    url = "https://www.contributor-covenant.org/version/3/0/code_of_conduct/",
    alt = "Contributor Covenant 3.0 Code of Conduct"
  )
)
```
<!-- badges: end -->

## Overview

Assessing and planning infrastructure and service networks, given a dispersed demand, limited capacity, accessibility targets, and concerns about spatial justice, is a central policy challenge. Problems of this type are commonly referred to as *Maximal Coverage Location-Allocation* (MCLA) spatial optimization problems.

`locationallocation` is an R package that provides tools for solving MCLA problems with geospatial data. It builds on widely used spatial libraries in R, follows [tidyverse principles](https://tidyverse.tidyverse.org/articles/manifesto.html), and integrates seamlessly with the broader [tidyverse ecosystem](https://tidyverse.org/). The package can generate travel-time maps and optimize the placement of facilities or infrastructure according to accessibility criteria, which can be weighted by one or more variables or by a user-defined function.

Potential applications of the package extend to the domains of public infrastructure assessment and planning (public services provision, e.g. transport, social services, healthcare, parks), urban environmental and climate risk reduction interventions, logistics and hubs allocation, commercial and strategic decisions.

> If you find this project useful, please consider giving it a star! &nbsp; [![GitHub Repository Stars](https://img.shields.io/github/stars/giacfalk/locationallocation)](https://github.com/giacfalk/locationallocation/)

## Installation

You can install `locationallocation` using the [`remotes`](https://github.com/r-lib/remotes) package:

```r
# install.packages(remotes)
remotes::install_github("giacfalk/locationallocation")
```

A [CRAN](https://cran.r-project.org/) version of the package is planned for the near future.

## Usage

To use the package, start by loading it to your R session using the `library` function:

```{r}
#| output: false

library(locationallocation)
```

As an example, we demonstrate how the package can address urban-scale climate risk through infrastructure assessment and geospatial planning. For this demonstration, we use the demo datasets included with the package.

These datasets include the coordinates of public drinking water fountains (blue dots) in Naples, Italy ([`naples_fountains`](https://giacfalk.github.io/locationallocation/reference/naples_fountains.html)); a gridded population layer provided as a `stars` raster from the Global Human Settlement Layer ([GHSL](https://human-settlement.emergency.copernicus.eu)) Population Grid ([GHS-POP](https://human-settlement.emergency.copernicus.eu/download.php?ds=pop)) ([`naples_population`](https://giacfalk.github.io/locationallocation/reference/naples_population.html)); a 100-meter resolution heat hazard map, representing the number of days with [Wet-Bulb Globe Temperature](https://en.wikipedia.org/wiki/Wet-bulb_globe_temperature) above 25°C during 2008–2017, obtained from the [UrbClim](https://www.urban-climate.eu/model) model ([`naples_hot_days`](https://giacfalk.github.io/locationallocation/reference/naples_hot_days.html)); and the city’s administrative boundaries ([`naples_shape`](https://giacfalk.github.io/locationallocation/reference/naples_shape.html)).

```{r}
#| label: existing-facilities
#| echo: false

library(ggplot2)
library(sf)
library(stars)

plot <-
  ggplot() +
  geom_raster(
    data = naples_population |> as.data.frame(xy = TRUE),
    mapping = aes(x = x, y = y, fill = pop_napoli),
  ) +
  geom_sf(
    data = naples_shape,
    fill = NA,
    color = "black"
  ) +
  geom_sf(
    data = naples_fountains |> st_filter(naples_shape),
    color = "blue",
    shape = 1,
    size = 2
  ) +
  scale_fill_terrain_c(
    na.value = "white"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Population"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank())

  plot |> locationallocation:::add_plot_annotation()
```

We can use the [`traveltime()`](https://giacfalk.github.io/locationallocation/reference/traveltime.html) function to create a map of current accessibility to the facility points (represented by the [`sf`](https://r-spatial.github.io/sf/) object [`naples_fountains`](https://giacfalk.github.io/locationallocation/reference/naples_fountains.html)) within the specified geographical boundaries. The function allows the user to select a travel mode (walking or fastest route) and an output spatial resolution in meters, which is achieved by resampling the friction layer with `stars`.

```{r}
#| output: false

traveltime_data <-
  naples_fountains |>
  traveltime(
    bb_area = naples_shape,
    dowscaling_model_type = "lm",
    mode = "walk",
    res_output = 100
  )
```

Once [`traveltime()`](https://giacfalk.github.io/locationallocation/reference/traveltime.html) has completed, the resulting layer can be visualized using the [`traveltime_plot()`](https://giacfalk.github.io/locationallocation/reference/traveltime_plot.html) function.

```{r}
#| label: traveltime-plot

traveltime_data |>
  traveltime_plot(
    bb_area = naples_shape,
    facilities = naples_fountains,
    contour_traveltime = 15
  )
```

We can also generate a summary plot and compute statistics using the output of the [`traveltime()`](https://giacfalk.github.io/locationallocation/reference/traveltime.html) function, together with a demand `stars` raster (e.g., population density) and a specified time threshold, using the [`traveltime_stats`](https://giacfalk.github.io/locationallocation/reference/traveltime_stats.html) function:

```{r}
#| label: traveltime-stats

traveltime_data |>
  traveltime_stats(
    demand = naples_population,
    breaks = c(5, 10, 15, 30),
    objectiveminutes = 15
  )
```

We can now use the [`allocation()`](https://giacfalk.github.io/locationallocation/reference/allocation.html) function to optimize the placement of new water fountains, ensuring that (virtually) everyone (i.e., the full extent of the `stars` layer specified by the `demand` parameter) can reach one within 15 minutes, as defined by the `objectiveminutes` parameter:

```{r}
#| output: false

allocation_data <-
  naples_population |>
  allocation(
    bb_area = naples_shape,
    facilities = naples_fountains,
    traveltime = traveltime_data,
    weights = NULL,
    objectiveminutes = 15,
    objectiveshare = 0.99,
    heur = "max",
    approach = "norm",
    exp_demand = 1,
    exp_weights = 1
  )
```

```{r}
#| echo: false

allocation_data |>
  locationallocation:::coverage_message()
```

```{r}
#| label: allocation-plot-1

allocation_data |> allocation_plot(naples_shape)
```

Note that it is also possible to solve an allocation problem using the `weights` parameter, which assigns greater relative importance or priority to areas where demand overlaps with weighting factors defined by another `stars` layer, such as exposure to hot days, as shown in the following example:

```{r}
#| output: false

allocation_data <-
  naples_population |>
  allocation(
    bb_area = naples_shape,
    facilities = naples_fountains,
    traveltime = traveltime_data,
    weights = naples_hot_days, # <--- Changed
    objectiveminutes = 15,
    objectiveshare = 0.99,
    heur = "max",
    approach = "norm",
    exp_demand = 1,
    exp_weights = 1
  )
```

```{r}
#| echo: false

allocation_data |>
  locationallocation:::coverage_message()
```

```{r}
#| label: allocation-plot-2

allocation_data |> allocation_plot(naples_shape)
```

It is also possible to apply normalization and exponentiation to different demand and weighting layers, enhancing their relative influence on the allocation, using the `approach`, `exp_demand`, and `exp_weights` parameters (see the [function documentation](https://giacfalk.github.io/locationallocation/reference/allocation.html) for details):

```{r}
#| output: false

allocation_data <-
  naples_population |>
  allocation(
    bb_area = naples_shape,
    facilities = naples_fountains,
    traveltime = traveltime_data,
    weights = naples_hot_days,
    objectiveminutes = 15,
    objectiveshare = 0.99,
    heur = "max",
    approach = "norm",
    exp_demand = 2, # <--- Changed
    exp_weights = 1
  )
```

```{r}
#| echo: false

allocation_data |>
  locationallocation:::coverage_message()
```

```{r}
#| label: allocation-plot-3

allocation_data |> allocation_plot(naples_shape)
```

A variant of the allocation problem arises when the set of candidate locations for new facilities is **discrete**, rather than **continuous** across the study area as in the previous example.

In this case, the user must provide a set of candidate points via the `candidate` parameter of the [`allocation_discrete()`](https://giacfalk.github.io/locationallocation/reference/allocation_discrete.html) function, along with the maximum number of facilities that can be selected (`n_fac` parameter). The function applies a quasi-optimality heuristic based on a randomization approach, where the number of replications (`n_samples` parameter) gradually approaches the global optimum, although computational time increases linearly. As with the continuous allocation problem, a weight layer and normalization and exponentiation parameters can also be specified.

```{r}
#| output: false

library(sf)

allocation_data <-
  naples_population |>
  allocation_discrete(
    bb_area = naples_shape,
    candidate = naples_shape |> st_sample(20),
    facilities = naples_fountains,
    n_fac = 5,
    n_samples = 100,
    traveltime = traveltime_data,
    weights = NULL,
    objectiveminutes = 15,
    objectiveshare = NULL,
    approach = "norm",
    exp_demand = 1,
    exp_weights = 1,
    par = FALSE
  )
```

```{r}
#| echo: false

allocation_data |>
  locationallocation:::coverage_message()
```

```{r}
#| label: allocation-plot-4

allocation_data |> allocation_plot(naples_shape)
```

Now consider a scenario where the user wants to choose up to `n_fac` facilities to meet a target `objectiveshare` of the total demand, assuming that level of coverage is actually attainable.

```{r}
#| output: false

library(sf)

allocation_data <-
  naples_population |>
  allocation_discrete(
    bb_area = naples_shape,
    candidate = naples_shape |> st_sample(20),
    facilities = naples_fountains,
    n_fac = 5,
    n_samples = 100,
    traveltime = traveltime_data,
    weights = NULL,
    objectiveminutes = 15,
    objectiveshare = 0.9,
    approach = "norm",
    exp_demand = 1,
    exp_weights = 1,
    par = FALSE
  )
```

```{r}
#| echo: false

allocation_data |>
  locationallocation:::coverage_message()
```

```{r}
#| label: allocation-plot-5

allocation_data |> allocation_plot(naples_shape)
```

Finally, consider a case with no preexisting facilities. The discrete location-allocation problem must then identify where to place new facilities to cover as much demand as possible, given the limited set of spatial options and the cap on how many facilities can be allocated.

```{r}
#| output: false

library(sf)

allocation_data <-
  naples_population |>
  allocation_discrete(
    bb_area = naples_shape,
    candidate = naples_shape |> st_sample(20),
    facilities = NULL,
    n_fac = 5,
    n_samples = 100,
    traveltime = NULL,
    weights = NULL,
    objectiveminutes = 15,
    objectiveshare = 0.45,
    approach = "norm",
    exp_demand = 1,
    exp_weights = 1,
    par = FALSE
  )
```

```{r}
#| echo: false

allocation_data |>
  locationallocation:::coverage_message()
```

```{r}
#| label: allocation-plot-6

allocation_data |> allocation_plot(naples_shape)
```

## Citation

If you use this package in your research, please cite it to acknowledge the effort put into its development and maintenance. Your citation helps support its continued improvement.

```{r}
citation("locationallocation")
```

## License

```{r}
#| echo: false
#| output: asis

cat(
  badge_license(
    license = "GPLv3",
    color = "bd0000",
    url = "https://www.gnu.org/licenses/gpl-3.0"
  )
)
```

```text
Copyright (C) 2025 Giacomo Falchetta

locationallocation is free software: you can redistribute it and/or modify it
under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any
later version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A
PARTICULAR PURPOSE. See the GNU General Public License for more details.

You should have received a copy of the GNU General Public License along with
this program. If not, see <https://www.gnu.org/licenses/>.
```

## Contributing

```{r}
#| echo: false
#| output: asis

cat(
  badge_custom(
    x = "Contributor%20Covenant",
    y = "3.0",
    color = "4baaaa",
    url = "https://www.contributor-covenant.org/version/3/0/code_of_conduct/",
    alt = "Contributor Covenant 3.0 Code of Conduct"
  )
)
```

Contributions are welcome! Whether you want to report bugs, suggest features, or improve the code or documentation, your input is highly valued. Please check the [issues tab](https://github.com/giacfalk/locationallocation/issues) for existing issues or to open a new one.
